
## Data preparation 
#### Dataset: 2020 NFL Predictions

```{r setup1, include = FALSE}
library(readxl)
library(tidyr)
library(readr)
library(ggplot2)
library(dplyr)
```

``` {r load_data, echo=FALSE, warning=FALSE}
sheet1 <- read_excel("/Users/Reeba/Downloads/ANA 515 data prep/515p/nfl_elo_latest.xlsx", sheet = "Sheet 1") 
sheet2 <- read_excel("/Users/Reeba/Downloads/ANA 515 data prep/515p/nfl_elo_latest.xlsx", sheet = "Sheet 2")

# Using dplyr for automatic column alignment
#combined_data <- bind_rows(sheet1, sheet2)

# got "Can't combine `..1$elo1_pre` <character> and `..2$elo1_pre` <double>" error. 
# Let's compare the datatypes and find the mismatched ones
```

#### Comparing data types between sheet 1 and sheet 2
```{r compare_data_types, echo=FALSE, warning=FALSE}
types_sheet1 <- sapply(sheet1, class)
types_sheet2 <- sapply(sheet2, class)

# Combine and compare data types
comparison <- data.frame(Column = names(types_sheet1),
                         Sheet1_Type = types_sheet1,
                         Sheet2_Type = types_sheet2[names(types_sheet1)])

mismatched_columns <- comparison[comparison$Sheet1_Type != comparison$Sheet2_Type, ]
mismatched_columns
```

#### Convert character data type to numeric data type and combine the sheets
```{r change_data_types_and_combine, echo=FALSE, warning=FALSE}
cols_to_convert <- c("elo1_pre", "elo2_post", "qb1_value_pre", "score1", "score2")
sheet1[cols_to_convert] <- lapply(sheet1[cols_to_convert], as.numeric)
sheet2[cols_to_convert] <- lapply(sheet2[cols_to_convert], as.numeric)

#str(sheet1)
#str(sheet2)

#datatypes are same, let's combine the sheets using the bind_rows function in the load_data chunk
combined_data <- bind_rows(sheet1, sheet2)
str(combined_data)
```

#### 1. The dates are in numeric format, let's convert them to date format.
```{r 1_date, echo=FALSE}
combined_data$date <- as.Date(combined_data$date, origin = "1899-12-30")
table(combined_data$date)
```
#### 2. Removing a row with an unusual date (1905-07-12) as it is irrelevant in a 2020 NFL prediction dataset
```{r 2_date, echo=FALSE}
combined_data <- combined_data[combined_data$date != as.Date("1905-07-12"), ]
table(combined_data$date)
```

#### 3. All values of 'season' should be 2020 since this prediction was done on a 2020 dataset. 
```{r season_before, echo=FALSE}
summary(combined_data$season)

#Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
# 2    2020    2020    1998    2020    2020 
```

Not all values are 2020. Lets correct them.
```{r season_after, echo=FALSE}
combined_data$season <- 2020
summary(combined_data$season)
# Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
# 2020    2020    2020    2020    2020    2020 
```

#### 4. Handing missing values in 'playoff' column and removing the 'neutral' column
Labeling them as NA so it is easy to analyze results
Also, let's flag and remove the 'neutral' column because except for 3 errored values (9,99,99) and others are 0, meaning they home/away games. 
```{r playoff, echo=FALSE, warning=FALSE}
# The empty values could either mean that the game was not a playoff or that the data was not recorded. Eitherway, it seems irrelevant for analysis.
combined_data$playoff[is.na(combined_data$playoff)] <- NA
print("Summary of playoff\n")
table(combined_data$playoff, useNA = "ifany")

neutral_column <- combined_data$neutral
# Remove the 'neutral' column from the dataset
combined_data$neutral <- NULL
```
#### 5.Handling missing values in the 'team 1' column
```{r team 1 before, echo=FALSE}
table(combined_data$team1, useNA = "ifany")
na_rows_team1 <- combined_data[is.na(combined_data$team1), ]
print(na_rows_team1[c("team1", "qb1")])
```
There are 4 rows with NA. Let's remove the row where all values are empty.
Let's find and impute the team names from the quaterbacks respective to these rows.
```{r team 1 after, echo=FALSE}
qb1_values <- na_rows_team1$qb1
# Kyler Murray - ARI
# Aaron Rodgers - GB
# John Wolford - LAR

# A function to update team1 based on qb1 values
for (i in 1:length(qb1_values)) {
  qb_name <- qb1_values[i]
  team <- combined_data$team1[combined_data$qb1 == qb_name]
  if (length(team) > 0) {
    combined_data$team1[is.na(combined_data$team1) & combined_data$qb1 == qb_name] <- team[1]
  }
}

#Two rows are imputed with the correct values. 
#The value of team1 where qb1 is John Wolford has to be imputed manually.
combined_data$team1[combined_data$qb1 == "John Wolford"] <- "LAR"
#Let's remove the remaining row where all the values of all columns are NA. 
combined_data <- combined_data[!is.na(combined_data$team1), ]
na_rows_team1 <- combined_data[is.na(combined_data$team1), ]
table(combined_data$team1, useNA = "ifany")
```
#### 6.Handling missing values in the 'team 2' column
```{r team 2 before, echo=FALSE}
table(combined_data$team2, useNA = "ifany")
na_rows_team2 <- combined_data[is.na(combined_data$team2), ]
print(na_rows_team2[c("team2", "qb2")])
```
1 - Three missing values, let's deal with them as we did for 'team 1' column
2 - Abbreviations of Houston, and oakland has to be corrected

```{r team 2 after, echo=FALSE}
qb2_values <- unique(na_rows_team2$qb2)

for (qb_name in qb2_values) {
  team_candidates <- combined_data$team2[combined_data$qb2 == qb_name & !is.na(combined_data$team2)]
  #cat("Found valid teams for", qb_name, ":", team_candidates, "\n")
  if (length(team_candidates) > 0) {
    # We can take the most common team from candidates if there are multiple
    common_team <- as.character(names(sort(table(team_candidates), decreasing = TRUE)[1])) # Most frequent team
    combined_data$team2[is.na(combined_data$team2) & combined_data$qb2 == qb_name] <- common_team
  } else {
    cat("No valid team found for", qb_name, "\n")
  }
}

combined_data$team2 <- dplyr::recode(combined_data$team2, "OAKLAND" = "OAK", "Houston" = "HOU")
table(combined_data$team2, useNA = "ifany")
```

#### 7. elo1_pre: The Elo rating for team 1 before the match. Let's handle outliers and missing values
```{r elo1_pre before, echo=FALSE,warning=FALSE}
summary(combined_data$elo1_pre)

#Min. 1st Qu.  Median    Mean 3rd Qu.    Max.    NA's 
#  1    1438    1515    1509    1590    1778       4 

# from the summary stats, it looks like the value 1 is an outlier as it is significantly below the quantiles, mean, and max values.
# Let's visualize using a boxplot to confirm it
ggplot(combined_data, aes(y=elo1_pre)) +
  geom_boxplot() +
  labs(title="Boxplot of elo1_pre", y="Values")
```

<br>
50% of the data falls within a range close to 1500 while the rest ranging from around 1200 to 1800.
There is one outlier around 0, represented by the dot. Let's impute the missing values and the outlier with the median value.
Since we have outliers, it is best to impute with median instead of mean
```{r elo1_pre after, echo=FALSE}
lower_bound <- 1200
flagged_rows_elo1_pre <- combined_data[is.na(combined_data$elo1_pre) | combined_data$elo1_pre < lower_bound, ]
#flagged_rows_elo1_pre
median_elo1_pre <- median(combined_data$elo1_pre, na.rm = TRUE)
combined_data$elo1_pre[combined_data$elo1_pre < 1200 | is.na(combined_data$elo1_pre)] <- median_elo1_pre
summary(combined_data$elo1_pre)
# Min. 1st Qu.  Median    Mean 3rd Qu.    Max. 
# 1277    1440    1515    1515    1588    1778 
ggplot(combined_data, aes(y=elo1_pre)) +
  geom_boxplot() +
  labs(title="Boxplot of elo1_pre", y="Values")
#the summary stats and the boxplot look good now.
```

#### 8. elo2_pre: The Elo rating for team 2 before the match. Doing the same as we did for elo1_pre
```{r elo2_pre before, echo=FALSE, warning=FALSE}
summary(combined_data$elo2_pre)

#Min. 1st Qu.  Median    Mean 3rd Qu.    Max.    NA's 
# 1260    1441    1512    1533    1590    8000       2 

ggplot(combined_data, aes(y=elo2_pre)) +
  geom_boxplot() +
  labs(title="Boxplot of elo2_pre", y="Values")
```

Its the opposite with this column, there is an outlier around 8000 and 2 missing values.
```{r elo2_pre after, echo=FALSE, warning=FALSE}
upper_bound <- 4000
flagged_rows_elo2_pre <- combined_data[is.na(combined_data$elo2_pre) | combined_data$elo2_pre > upper_bound, ]
#flagged_rows_elo2_pre
median_elo2_pre <- median(combined_data$elo2_pre, na.rm = TRUE)
combined_data$elo2_pre[combined_data$elo2_pre > 4000 | is.na(combined_data$elo2_pre)] <- median_elo2_pre
summary(combined_data$elo2_pre)
ggplot(combined_data, aes(y=elo2_pre)) +
  geom_boxplot() +
  labs(title="Boxplot of elo2_pre", y="Values")
```


#### 9. Handling outliers and missing values in columns elo_prob1, and elo_prob2
```{r elo_prob before, echo=FALSE}
columns <- c("elo_prob1", "elo_prob2")
for (col in columns) {
  cat("\nSummary for", col, ":\n")
  print(summary(combined_data[[col]]))  
}
#ggplot(combined_data, aes(y=combined_data$elo_prob1)) + geom_boxplot() + labs(title="Boxplot of elo_prob1", y="Values")
#ggplot(combined_data, aes(y=combined_data$elo_prob2)) + geom_boxplot() + labs(title="Boxplot of elo_prob2", y="Values")
```
Upper bound outlier found in elo_prob2 (Max - 4000).
Replace values of elo_prob2 with (1-elo_prob1) according to the formula
```{r elo_prob after, echo=FALSE}
flagged_rows_elo_prob1 <- combined_data[is.na(combined_data$elo_prob1), ]
flagged_rows_elo_prob2 <- combined_data[is.na(combined_data$elo_prob2), ]

median_elo_prob1 <- median(combined_data$elo_prob1, na.rm = TRUE)
combined_data$elo_prob1[is.na(combined_data$elo_prob1)] <- median_elo_prob1
combined_data$elo_prob2 <- 1 - combined_data$elo_prob1

for (col in columns) {
  cat("\nSummary for", col, ":\n")
  print(summary(combined_data[[col]]))  
}
```
#### 10. Handling negative outliers and missing values in columns elo1_post, and elo2_post
```{r elo_post before, echo=FALSE, warning=FALSE}
columns <- c("elo1_post", "elo2_post")
for (col in columns) {
  cat("\nSummary for", col, ":\n")
  print(summary(combined_data[[col]]))  
}

#ggplot(combined_data, aes(y=elo1_post)) + geom_boxplot() + labs(title="Boxplot of elo1_post", y="Values")
#ggplot(combined_data, aes(y=elo2_post)) + geom_boxplot() + labs(title="Boxplot of elo2_post", y="Values")
```
elo1_post - outlier at 0, it should be the value 1 (min) and two missing values.
elo2_post - negative outlier, elo ratings are never negative. Let's impute it with median

```{r elo_post after, echo=FALSE, warning=FALSE}
lower_bound <- 1000
flagged_rows_elo1_post <- combined_data[is.na(combined_data$elo1_post) | combined_data$elo1_post < lower_bound, ]
#flagged_rows_elo1_post
median_elo1_post <- median(combined_data$elo1_post, na.rm = TRUE)
combined_data$elo1_post[combined_data$elo1_post < 1000 | is.na(combined_data$elo1_post)] <- median_elo1_post

lower_bound <- 0
flagged_rows_elo2_post <- combined_data[is.na(combined_data$elo2_post) | combined_data$elo2_post < lower_bound, ]
#flagged_rows_elo2_post
# 7 missing values and some negative values (outliers)
median_elo2_post <- median(combined_data$elo2_post, na.rm = TRUE)
combined_data$elo2_post[combined_data$elo2_post < 0 | is.na(combined_data$elo2_post)] <- median_elo2_post

for (col in columns) {
  cat("\nSummary for", col, ":\n")
  print(summary(combined_data[[col]]))  
}

```
#### 11. Handling outliers and missing values in columns qbelo1_pre, and qbelo2_pre
```{r qbelo_pre before, echo=FALSE, warning=FALSE}
columns <- c("qbelo1_pre", "qbelo2_pre")
for (col in columns) {
  cat("\nSummary for", col, ":\n")
  print(summary(combined_data[[col]]))  
}

#ggplot(combined_data, aes(y=qbelo1_pre)) + geom_boxplot() + labs(title="Boxplot of qbelo1_pre", y="Values")
#ggplot(combined_data, aes(y=qbelo2_pre)) + geom_boxplot() + labs(title="Boxplot of qbelo2_pre", y="Values")
```

qbelo1_pre - Quarterback Elo ratings, a value of 0 is highly unlikely and unrealistic. It looks like an error, so, lets impute them with median
qbelo2_pre - Missing values 
```{r qbelo_pre after, echo=FALSE, warning=FALSE}
lower_bound <- 1000
flagged_rows_qbelo1_pre <- combined_data[is.na(combined_data$qbelo1_pre) | combined_data$qbelo1_pre < lower_bound, ]
flagged_rows_qbelo2_pre <- combined_data[is.na(combined_data$qbelo2_pre), ]

median_qbelo1_pre <- median(combined_data$qbelo1_pre, na.rm = TRUE)
combined_data$qbelo1_pre[combined_data$qbelo1_pre < lower_bound | is.na(combined_data$qbelo1_pre)] <- median_qbelo1_pre

median_qbelo2_pre <- median(combined_data$qbelo2_pre, na.rm = TRUE)
combined_data$qbelo2_pre[is.na(combined_data$qbelo2_pre)] <- median_qbelo2_pre

for (col in columns) {
  cat("\nSummary for", col, ":\n")
  print(summary(combined_data[[col]]))  
}
```
#### 12. Handling missing values in columns qb1, and qb2
```{r qb before, echo=FALSE, warning=FALSE}
#table(combined_data$qb1, useNA = "ifany")
#table(combined_data$qb2, useNA = "ifany")

na_rows_qb1 <- combined_data[is.na(combined_data$qb1), ]
print(na_rows_qb1[c("qb1", "team1")])
na_rows_qb2 <- combined_data[is.na(combined_data$qb2), ]
print(na_rows_qb2[c("qb2", "team2")])
```
Let's find and update the quaterback names from the teams we found and remove the row where all values are NA.
Also, scan for spelling mistakes or special characters. None found
```{r qb after, echo=FALSE}
#find the team1 values of the empty qb1 rows
team1_values <- na_rows_qb1$team1
team2_values <- na_rows_qb2$team2


# Update qb1 based on team1's qb1 values
for (i in 1:length(team1_values)) {
  team1_name <- team1_values[i]
  qb <- combined_data$qb1[combined_data$team1 == team1_name]
  if (length(qb) > 0) {
    combined_data$qb1[is.na(combined_data$qb1) & combined_data$team1 == team1_name] <- qb[1]
  }
}
# Update qb2 based on team2's qb2 value
for (i in 1:length(team2_values)) {
  team2_name <- team2_values[i]
  qb_2 <- combined_data$qb2[combined_data$team2 == team2_name]
  if (length(qb_2) > 0) {
    combined_data$qb2[is.na(combined_data$qb2) & combined_data$team2 == team2_name] <- qb_2[1]
  }
}

combined_data <- combined_data[!is.na(combined_data$qb1), ]

table(combined_data$qb1, useNA = "ifany")
table(combined_data$qb2, useNA = "ifany")
```

#### 13. Handling missing values in columns qb1_value_pre, and qb2_value_pre
```{r qb_value_pre before, echo=FALSE, warning=FALSE}
#summary(combined_data$qb2_value_pre)
#its data type is set as character because of an NA value, let's convert it to numeric
combined_data$qb2_value_pre <- as.numeric(combined_data$qb2_value_pre)
columns <- c("qb1_value_pre", "qb2_value_pre")
for (col in columns) {
  cat("\nSummary for", col, ":\n")
  print(summary(combined_data[[col]]))  
}
ggplot(combined_data, aes(y=qb1_value_pre)) + geom_boxplot() + labs(title="Boxplot of qb1_value_pre", y="Values")
ggplot(combined_data, aes(y=qb2_value_pre)) + geom_boxplot() + labs(title="Boxplot of qb2_value_pre", y="Values")
```


potential outliers around 0, a realistic minimum value for a quarterback's Elo rating in established leagues should be above 100 or 150, as Elo ratings typically start around 1500.
```{r qb_value_pre after, echo=FALSE}
lower_bound = 100
flagged_rows_qb1_value_pre <- combined_data[is.na(combined_data$qb1_value_pre) | combined_data$qb1_value_pre < lower_bound, ]
flagged_rows_qb2_value_pre <- combined_data[is.na(combined_data$qb2_value_pre) | combined_data$qb2_value_pre < lower_bound, ]

median_qb1_value_pre <- median(combined_data$qb1_value_pre, na.rm = TRUE)
combined_data$qb1_value_pre[is.na(combined_data$qb1_value_pre)| combined_data$qb1_value_pre < lower_bound] <- median_qb1_value_pre
median_qb2_value_pre  <- median(combined_data$qb2_value_pre , na.rm = TRUE)
combined_data$qb2_value_pre[is.na(combined_data$qb2_value_pre) | combined_data$qb2_value_pre < lower_bound ] <- median_qb2_value_pre 
for (col in columns) {
  cat("\nSummary for", col, ":\n")
  print(summary(combined_data[[col]]))  
}

```
#### 14. Handling missing values in columns qb1_adj, and qb2_adj
```{r qb_adj before, echo=FALSE, warning=FALSE}
combined_data$qb1_adj <- as.numeric(combined_data$qb1_adj)
columns <- c("qb1_adj", "qb2_adj")
for (col in columns) {
  cat("\nSummary for", col, ":\n")
  print(summary(combined_data[[col]]))  
}

#ggplot(combined_data, aes(y=qb1_adj)) + geom_boxplot() + labs(title="Boxplot of qb1_adj", y="Values")
#ggplot(combined_data, aes(y=qb2_adj)) + geom_boxplot() +labs(title="Boxplot of qb2_adj", y="Values")
```

Both the adjusted values have outliers and NAs. let's find the lower and upper bounds by calculating IQR.
And, replace the outliers < lower bound | outliers > upper bound | NAs with median
```{r qb_adj after, echo=FALSE}
# Step 1: a function to calculate the IQR, lower and upper bounds
calculate_bounds <- function(x) {
  Q1 <- quantile(x, 0.25, na.rm = TRUE)
  Q3 <- quantile(x, 0.75, na.rm = TRUE)
  IQR <- Q3 - Q1
  lower_bound <- Q1 - 1.5 * IQR
  upper_bound <- Q3 + 1.5 * IQR
  return(list(lower_bound = lower_bound, upper_bound = upper_bound))
}

# Step 2: Calculate the bounds for qb1_adj and qb2_adj
bounds_qb1_adj <- calculate_bounds(combined_data$qb1_adj)
lower_bound_qb1_adj <- bounds_qb1_adj$lower_bound
upper_bound_qb1_adj <- bounds_qb1_adj$upper_bound

bounds_qb2_adj <- calculate_bounds(combined_data$qb2_adj)
lower_bound_qb2_adj <- bounds_qb2_adj$lower_bound
upper_bound_qb2_adj <- bounds_qb2_adj$upper_bound

flagged_rows_qb1_adj <- combined_data[is.na(combined_data$qb1_adj) | combined_data$qb1_adj > upper_bound_qb1_adj | combined_data$qb1_adj < lower_bound_qb1_adj,]
flagged_rows_qb2_adj <- combined_data[is.na(combined_data$qb2_adj) | combined_data$qb2_adj > upper_bound_qb2_adj | combined_data$qb2_adj < lower_bound_qb2_adj,]

# Step 3: Calculate the median for qb1_adj and qb2_adj
median_qb1_adj <- median(combined_data$qb1_adj, na.rm = TRUE)
combined_data$qb1_adj[is.na(combined_data$qb1_adj) | combined_data$qb1_adj > upper_bound_qb1_adj | combined_data$qb1_adj < lower_bound_qb1_adj] <- median_qb1_adj

median_qb2_adj <- median(combined_data$qb2_adj, na.rm = TRUE)
combined_data$qb2_adj[is.na(combined_data$qb2_adj) | combined_data$qb2_adj > upper_bound_qb2_adj | combined_data$qb2_adj < lower_bound_qb2_adj] <- median_qb2_adj
for (col in columns) {
  cat("\nSummary for", col, ":\n")
  print(summary(combined_data[[col]]))  
}

```


#### 15. Handling missing values in columns qbelo_prob1, and qbelo_prob2
```{r qbelo_prob before, echo=FALSE, warning=FALSE}
columns <- c("qbelo_prob1", "qbelo_prob2")
for (col in columns) {
  cat("\nSummary for", col, ":\n")
  print(summary(combined_data[[col]]))  
}
#ggplot(combined_data, aes(y=qbelo_prob1)) + geom_boxplot() +labs(title="Boxplot of qbelo_prob1", y="Values")
#ggplot(combined_data, aes(y=qbelo_prob2)) +geom_boxplot() +labs(title="Boxplot of qbelo_prob2", y="Values")
```

Both have 3 NAs each, qbelo_prob1 alone has an outlier (25)

```{r qbelo_prob after, echo=FALSE}
upper_bound = 20
flagged_rows_qbelo_prob1 <- combined_data[is.na(combined_data$qbelo_prob1)| combined_data$qbelo_prob1 > upper_bound,]
flagged_rows_qbelo_prob2 <- combined_data[is.na(combined_data$qbelo_prob2), ]

median_qbelo_prob1 <- median(combined_data$qbelo_prob1, na.rm = TRUE)
combined_data$qbelo_prob1[is.na(combined_data$qbelo_prob1)| combined_data$qbelo_prob1 > upper_bound] <- median_qbelo_prob1

median_qbelo_prob2 <- median(combined_data$qbelo_prob2, na.rm = TRUE)
combined_data$qbelo_prob2[is.na(combined_data$qbelo_prob2)] <- median_qbelo_prob2
for (col in columns) {
  cat("\nSummary for", col, ":\n")
  print(summary(combined_data[[col]]))  
}
```
#### 16. Handling missing values in columns qb1_game_value, and qb2_game_value
```{r qb_game_value before, echo=FALSE, warning=FALSE}
columns <- c("qb1_game_value", "qb2_game_value")
for (col in columns) {
  cat("\nSummary for", col, ":\n")
  print(summary(combined_data[[col]]))  
}
ggplot(combined_data, aes(y=qb1_game_value)) + geom_boxplot() +labs(title="Boxplot of qb1_game_value", y="Values")
ggplot(combined_data, aes(y=qb2_game_value)) +geom_boxplot() +labs(title="Boxplot of qb2_game_value", y="Values")
```

Both have positive and negative outliers. Let's call the calculate_bounds function to find the upper and lower bounds, then replace them with median
```{r qb_game_value after, echo=FALSE, warning=FALSE}
bounds_qb1_game_value <- calculate_bounds(combined_data$qb1_game_value)
lower_bound_qb1_game_value <- bounds_qb1_game_value$lower_bound
upper_bound_qb1_game_value <- bounds_qb1_game_value$upper_bound

bounds_qb2_game_value <- calculate_bounds(combined_data$qb2_game_value)
lower_bound_qb2_game_value <- bounds_qb2_game_value$lower_bound
upper_bound_qb2_game_value <- bounds_qb2_game_value$upper_bound

flagged_rows_qb1_game_value <- combined_data[is.na(combined_data$qb1_game_value)| combined_data$qb1_game_value > upper_bound_qb1_game_value | combined_data$qb1_game_value < lower_bound_qb1_game_value,]
flagged_rows_qb2_game_value <- combined_data[is.na(combined_data$qb2_game_value)| combined_data$qb2_game_value > upper_bound_qb2_game_value | combined_data$qb2_game_value < lower_bound_qb2_game_value,]

median_qb1_game_value <- median(combined_data$qb1_game_value, na.rm = TRUE)
combined_data$qb1_game_value[is.na(combined_data$qb1_game_value)| combined_data$qb1_game_value > upper_bound_qb1_game_value | combined_data$qb1_game_value < lower_bound_qb1_game_value] <- median_qb1_game_value

median_qb2_game_value <- median(combined_data$qb2_game_value, na.rm = TRUE)
combined_data$qb2_game_value[is.na(combined_data$qb2_game_value)| combined_data$qb2_game_value > upper_bound_qb2_game_value | combined_data$qb2_game_value < lower_bound_qb2_game_value] <- median_qb2_game_value

for (col in columns) {
  cat("\nSummary for", col, ":\n")
  print(summary(combined_data[[col]]))  
}
ggplot(combined_data, aes(y=qb1_game_value)) + geom_boxplot() +labs(title="Boxplot of qb1_game_value", y="Values")
ggplot(combined_data, aes(y=qb2_game_value)) +geom_boxplot() +labs(title="Boxplot of qb2_game_value", y="Values")
```
In prediction models, especially for sports like NFL, outliers may represent extreme scenarios, such as unusually bad or good performances predicted for a quarterback. These could be rare but realistic outcomes. If the outliers are natural and don't heavily impact your analysis, it’s perfectly acceptable to leave them in the dataset.

#### 17. Handling missing values in columns qb1_game_value, qb2_game_value, qbelo1_post, and qbelo2_post
```{r qb_value_post qbelo_post before, echo=FALSE, warning=FALSE}

columns <- c("qb1_value_post", "qb2_value_post", "qbelo1_post", "qbelo2_post")
for (col in columns) {
  cat("\nSummary for", col, ":\n")
  print(summary(combined_data[[col]]))  
}

#ggplot(combined_data, aes(y=qb1_value_post)) + geom_boxplot() + labs(title="Boxplot of qb1_value_post", y="Values")
#ggplot(combined_data, aes(y=qb2_value_post)) + geom_boxplot() + labs(title="Boxplot of qb2_value_post", y="Values")
#ggplot(combined_data, aes(y=qbelo1_post)) + geom_boxplot() + labs(title="Boxplot of qbelo1_post", y="Values")
#ggplot(combined_data, aes(y=qbelo2_post)) + geom_boxplot() + labs(title="Boxplot of qbelo2_post", y="Values")
```
qb1_value_post - negative outlier
qbelo1_post - outlier around 0
```{r qb_value_post qbelo_post after, echo=FALSE, warning=FALSE}
flagged_rows_qb1_value_post <- combined_data[is.na(combined_data$qb1_value_post)| combined_data$qb1_value_post < 0,]
flagged_rows_qb2_value_post <- combined_data[is.na(combined_data$qb2_value_post),]
flagged_rows_qbelo1_post <- combined_data[is.na(combined_data$qbelo1_post)| combined_data$qbelo1_post < 20,]
flagged_rows_qbelo2_post <- combined_data[is.na(combined_data$qbelo2_post),]

median_qb1_value_post<- median(combined_data$qb1_value_post, na.rm = TRUE)
combined_data$qb1_value_post[is.na(combined_data$qb1_value_post)| combined_data$qb1_game_value < 0] <- median_qb1_value_post
median_qb2_value_post <- median(combined_data$qb2_value_post, na.rm = TRUE)
combined_data$qb2_value_post[is.na(combined_data$qb2_value_post)] <- median_qb2_value_post
median_qbelo1_post<- median(combined_data$qbelo1_post, na.rm = TRUE)
combined_data$qbelo1_post[is.na(combined_data$qbelo1_post)| combined_data$qbelo1_post < 20] <- median_qbelo1_post
median_qbelo2_post <- median(combined_data$qbelo2_post, na.rm = TRUE)
combined_data$qbelo2_post[is.na(combined_data$qbelo2_post)] <- median_qbelo2_post
for (col in columns) {
  cat("\nSummary for", col, ":\n")
  print(summary(combined_data[[col]]))
}
```

#### 18. Handling missing values in columns score 1 and score 2
```{r score before, echo=FALSE, warning=FALSE}

columns <- c("score1", "score2")
for (col in columns) {
  cat("\nSummary for", col, ":\n")
  print(summary(combined_data[[col]]))  
}

#ggplot(combined_data, aes(y=score1)) + geom_boxplot() + labs(title="Boxplot of score 1", y="Values")
#ggplot(combined_data, aes(y=score2)) + geom_boxplot() + labs(title="Boxplot of score 2", y="Values")

```
Score 1 - Negative scores are considered outliers as they are unrealistic, clearly errors. 
Also, a score of 99 is extremely unlikely

```{r score after, echo=FALSE, warning=FALSE}
flagged_rows_score1 <- combined_data[is.na(combined_data$score1) | combined_data$score1 < 0 | combined_data$score1 > 75,]
median_score1<- median(combined_data$score1, na.rm = TRUE)
combined_data$score1[is.na(combined_data$score1)| combined_data$score1 < 0 | combined_data$score1 > 75] <- median_score1

flagged_rows_score2 <- combined_data[is.na(combined_data$score2),]
median_score2<- median(combined_data$score2, na.rm = TRUE)
combined_data$score2[is.na(combined_data$score2)] <- median_score2

for (col in columns) {
  cat("\nSummary for", col, ":\n")
  print(summary(combined_data[[col]]))  
}
```


```{r export, echo=FALSE}
output_file <- "cleaned_dataset.csv"
write.csv(combined_data, file = output_file, row.names = FALSE)
cat("The cleaned dataset has been saved as:", output_file)
```

